---
title: "R oefeningen Inleiding R cursus"
author: "Longhow Lam"
output: 
  html_notebook:
    theme: sandstone
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

# Avond sessie 01

---

## Data types
A.	Create the object x1 with the numbers 1, 3.51, 0.7 and 5.
B.	Show the data class of x1.
C.	Transform x1 into an integer data type and store the result in x2.
D.	Use the round function to round x1 and store the result in x3.
E.	Is x3 of data type integer?


```{r}
x1 = c(1, 3.51, 0.7, 5)
class(x1)
x2 = as.integer(x1)
x3 = round(x1)
is.integer(x3)
```

## Logical data type.
A.	Create the vector y1 with the numbers 1, 2, 3, 4, 5, 6, 7, 8, 9 and 10.
B.	Create a logical vector y2 from y1. An element of y2 should be TRUE if the corresponding element of y1 is larger than 5.
C.	How many elements are smaller than 5?

```{r}
y1 = 1:10
y2 = y1 > 5
sum(y2)
```


## Character data type and factor data type.
A.	Create the character vector y1, with the characters “high”, “low”, “high”, “high”, “low”.
B.	Create the factor variable y2 from y1. Print the levels of y2.
C.	Transform y2 to a numeric vector y3, so that high becomes 2 and low becomes 1.

```{r}
y1 = c("high", "low", "high", "high", "low")
y2 = as.factor(y1)
levels(y2)

y2 = ordered(y2,levels = c("low","high") )
as.integer(y2)

```

## Creating sequences.
A.	Create a vector x, with elements (1,2,3,4,..,100).
B.	Create a vector y, with elements (0,5,10,...,500).
C.	Create a vector z1, with elements (1,1,1,2,2,2,...,50,50,50).
D.	Create a vector z2, with elements (1,2,2,3,3,3,.....,10,....10).
E.	Create a vector z3, with elements (1,2,2,3,4,4,5,6,6,7,8,8,...,50,50).

```{r}
x = 1:100
y = seq(from = 0, to = 500, by =5)
z1 = rep(1:50, each=3)
z2 = rep(1:10, 1:10)
z3 = rep(1:10, rep(1:2,5))
```

## Creating factors from double data.
A.	Generate a vector x with 100 random normal numbers.
B.	Use x to generate a factor vector y with three levels: “low”, “average” and “high”. If an element in x is:

*	between -Inf and -1.5 then the level in y should be “low”,
*	between -1.5 and 1.5 then the level in y should be “average”,
*	between 1.5 and Inf the level in y should be “high”.

Use the function `cut`.

```{r}
x = rnorm(100)
y = cut(x, breaks = c(-Inf,-1.5,1.5,Inf))
levels(y) = c("low", "average", "high")
x
y

```

## Creating a data.frame
A.	Create a data frame `df1` from the two vectors x and y from the previous exercise

B.	Rename the columns x and y into index and income

C.	Create an additional factor column Gender, randomly filled with “M” or “F”

```{r}
df1 = data.frame(x,y)
names(df1) = c("index", "income")
df1$Gender = as.factor(
  sample(
    c("M","F"), 
    replace = TRUE,
    size = dim(df1)[1]
    )
)
  
```

## Creating a list 

A.	Create a list L1 with 4 elements:

*	X1 = 100 random normal numbers,
*	X2  = 100 random uniform numbers between 0 and 1
*	X3 = the numbers one to 20
*	x4 = the numbers 50 to 1

B.  Give an overview of the components of L1

C. Calculate the mean of the four vectors in list L1

```{r}
L1 = list(
  X1 = rnorm(100),
  X2 = runif(200),
  X3 = 1:20,
  x5 = 50:1
)

names(L1)

lapply(L1, mean)
```

## Importing some data

In the data directory there are some dummy text files: f1.csv, f2.txt and f3.txt
Import these sets in the ‘correct’ way. The resulting datasets in R should all have 3 columns.

```{r}
f1 <- read_csv("C:/Users/llam/OneDrive - RTL Nederland B.V/Personal/inleidingR/data/f1.csv",  skip = 3)

f2 <- read_delim("C:/Users/llam/OneDrive - RTL Nederland B.V/Personal/inleidingR/data/f2.txt",   " ")

f3 <- read.csv("C:/Users/llam/OneDrive - RTL Nederland B.V/Personal/inleidingR/data/f3.txt")
```

<br>

# Avond sessie 02

---

## String bewerkingen
Copy the following vector to R.
X = c("abd 06-456", "blab la (06)-123.45678", "hoi 06-12123478", "mijn nr 0689452312")
A.	For this exercise a valid mobile phone number starts with 06 followed by hyphen or space or not and then 8 consecutive digits. What are the valid numbers in x, use a regular expression.
B.	Extract the phone numbers from x into a new vector y

```{r}
library(stringr)
x = c("abd 06-456", "blab la (06)-123.45678", "hoi 06-12123478", "mijn nr 0689452312")

mob_pattern = "[0][6][[:punct:]]*[\\d]{8}"
str_detect(x,mob_pattern)

str_extract(x, mob_pattern)
```

## Dates and times.
A.	Create a date object from the characters "31-jan-05"   "22-Sep-07".
B.	Create a date object from the characters "31-jan-05 3:03:20"  "22-Sep-07 22:29:56".
C.	Add 3 months and 2 days to the dates from exercise (B)
D.	Create a date sequence vector x with the dates: 31-1-2016, 28-2-2016, 31-03-2016,……,31-12-2016 

```{r}

```

For the following two exercises we are going to use restaurant data in the file Restaurants.csv in the data folder. Use the read_csv function to import this data set.

## Basic data manipulation 1
A.	Create a subset of the restaurants whose name is “kota radja”, how many are there?
B.	Give a top ten restaurants in Amsterdam based on price.
C.	Create an extra column that has the values zero or one.   1 if Kitchen type is Chinees and 0 otherwise.
D.	Create a sub set of the data, filter only the restaurants that have price smaller than 25.
E.	Give the number per kitchen types of that sub set.

```{r}

```

## Basic data manipulation 2
A.	In the data folder there is a Postcide_NL.RDs file, import that set. 
B.	Use it to enrich the restaurant data set whith an extra column provincie.
C.	Count the number of restaurants per province of the whole set.
D.	Give the max and min prices per province.
E.	In the column “keuken” there are some mistakes. A price where there should have been a kitchen type. Remove all kitchen types that have a digit in.

```{r}

```

## Auto data van gaspedaal
In the data folder there is a file R data set AllCarsGasPedaal.Rds  import this data set using readRDS

A.	List the names of the columns of the data set.
B.	In this data set there is a column Kilometerstand which is character, convert it to numeric so that we can calculate with it.
C.	Select only the brands Audi, Renault and BMW put those records in a new data frame.

## Split data and stack data
Consider the following simple data set with 3 rows:

Df = data.frame(
  V1 = c(“a,b,r”, “p,q”, “p,q,w,z”)
)

I need the following set:
Id Item
1 a
1 b
1 r
2 p
2 q
3 p
3 q
3 w
3 z
How to this? use tidyr functions separate and gather.

```{r}
library(tidyverse)
Df = data.frame(
  V1 = c("a,b,r", "p,q", "p,q,w,z")
)

tmp = tidyr::separate(Df, V1, into = paste0("V",1:10)) %>%
  mutate(tmp = 1:length(V1)) %>%
  gather(id, item, -tmp) %>%
  filter(!is.na(item))
```

## sampling van data 
Kijk naar sample_n of sample_frac uit dplyr.
A. Create a df1 data frame with 100 columns and 1000 rows. Filled with random numbers between 0 and 1.
B. Create df2 which samples 500 rows from df1
C. Create df3 which samples 50% of the rows from df1

```{r}

```



<br>

# Avond sessie 03

---

Use the set AllCarsGasPedaal.Rds  from the previous exercise.
```{r}
library(ggplot2)
library(dplyr)

AllCars = readRDS("data/AllCarsGasPedaal.Rds")
```

## plotting using ggplot2

A.	Create a scatterplot of the VraagPrijs and KMStand.
B.	Remove outlying points and create plot again using alpha = 0.05
C.  Add a smoothed line trough the points to get an idea of the relation between KM and Price.
D.	Filter the set so that we only have the top eight brands (based on # observations)
E.	Create a new scatter plot, now the points should be colored  by brand, using the subset of the previous exercise 
F.  Create also a smoothing line per brand
G.	Create facets by brand, each facet should have a histogram of the price, create also a grid of 2 rows and 4 columns. 
H.	Create facets by brand, each facet should contain a scatter plot between price and KM stand


```{r}
AllCars %>% 
  ggplot(aes(x=KM, y = Prijs)) +
  geom_point()

AllCars %>% 
  dplyr::filter(Prijs < 100000, KM < 500000) %>%
  ggplot(aes(x=KM, y = Prijs)) +
  geom_point(alpha = 0.05)

AllCars %>%
  dplyr::filter(Prijs < 100000, KM < 500000) %>%
  ggplot(aes(x=KM, y = Prijs)) +
  geom_point(alpha = 0.05) + 
  geom_smooth()

top8 = AllCars %>% group_by(Merk) %>% summarise(n=n()) %>% arrange(desc(n)) %>% slice(1:8) %>% select(Merk)

AllCars %>% 
  dplyr::filter(Prijs < 100000, KM < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(x=KM, y = Prijs)) +
  geom_point(aes(colour=Merk),alpha = 0.05) 

AllCars %>% 
  dplyr::filter(Prijs < 100000, KM < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(x=KM, y = Prijs)) +
  geom_point(aes(colour=Merk),alpha = 0.05) + 
  geom_smooth(aes(colour=Merk))

AllCars %>% 
  dplyr::filter(Prijs < 100000, KM < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(Prijs)) +
  geom_histogram() + facet_grid(~Merk)

AllCars %>% 
  dplyr::filter(Prijs < 100000, KM < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(Prijs)) +
  geom_histogram() + facet_wrap(~Merk, nrow=2)

AllCars %>% 
  dplyr::filter(Prijs < 100000, KM < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(x= KM, y = Prijs)) +
  geom_point() + facet_grid(~Merk)


```


## plotting using plotly


## plotting using leaflet
