---
title: "R oefeningen Inleiding R cursus"
author: "Longhow Lam"
output: 
  html_notebook:
    theme: sandstone
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

# Session 01. Data types & structures

---

## Data types
A.	Create the object x1 with the numbers 1, 3.51, 0.7 and 5. <br>
B.	Show the data class of x1. <br>
C.	Transform x1 into an integer data type and store the result in x2. <br>
D.	Use the round function to round x1 and store the result in x3. <br>
E.	Is x3 of data type integer?


```{r}
x1 = c(1, 3.51, 0.7, 5)
class(x1)
x2 = as.integer(x1)
x3 = round(x1)
is.integer(x3)
```

## Logical data type.
A.	Create the vector y1 with the numbers 1, 2, 3, 4, 5, 6, 7, 8, 9 and 10. <br>
B.	Create a logical vector y2 from y1. An element of y2 should be TRUE if the corresponding element of y1 is larger than 5. <br>
C.	How many elements are smaller than 5?

```{r}
y1 = 1:10
y2 = y1 > 5
sum(y2)
```


## Character data type and factor data type.
A.	Create the character vector y1, with the characters “high”, “low”, “high”, “high”, “low”. <br>
B.	Create the factor variable y2 from y1. Print the levels of y2. <br>
C.	Transform y2 to a numeric vector y3, so that high becomes 2 and low becomes 1. <br>

```{r}
y1 = c("high", "low", "high", "high", "low")
y2 = as.factor(y1)
levels(y2)

y2 = ordered(y2,levels = c("low","high") )
as.integer(y2)

```

## Creating sequences.
A.	Create a vector x, with elements (1,2,3,4,..,100). <br>
B.	Create a vector y, with elements (0,5,10,...,500). <br>
C.	Create a vector z1, with elements (1,1,1,2,2,2,...,50,50,50). <br>
D.	Create a vector z2, with elements (1,2,2,3,3,3,.....,10,....10). <br>
E.	Create a vector z3, with elements (1,2,2,3,4,4,5,6,6,7,8,8,...,50,50). <br>

```{r}
x = 1:100
y = seq(from = 0, to = 500, by =5)
z1 = rep(1:50, each=3)
z2 = rep(1:10, 1:10)
z3 = rep(1:10, rep(1:2,5))
```

## Creating factors from double data.
A.	Generate a vector x with 100 random normal numbers.  <br>
B.	Use x to generate a factor vector y with three levels: “low”, “average” and “high”. If an element in x is:  <br>

*	between -Inf and -1.5 then the level in y should be “low”,
*	between -1.5 and 1.5 then the level in y should be “average”,
*	between 1.5 and Inf the level in y should be “high”.

Use the function `cut`.

```{r}
x = rnorm(100)
y = cut(x, breaks = c(-Inf,-1.5,1.5,Inf))
levels(y) = c("low", "average", "high")
x
y

```

## Creating a data.frame
A.	Create a data frame `df1` from the two vectors x and y from the previous exercise

B.	Rename the columns x and y into index and income

C.	Create an additional factor column Gender, randomly filled with “M” or “F”

```{r}
df1 = data.frame(x,y)
names(df1) = c("index", "income")
df1$Gender = as.factor(
  sample(
    c("M","F"), 
    replace = TRUE,
    size = dim(df1)[1]
    )
)
  
```

## Creating a list 

A.	Create a list L1 with 4 elements:

*	X1 = 100 random normal numbers,
*	X2  = 100 random uniform numbers between 0 and 1
*	X3 = the numbers one to 20
*	x4 = the numbers 50 to 1

B.  Give an overview of the components of L1

C. Calculate the mean of the four vectors in list L1

```{r}
L1 = list(
  X1 = rnorm(100),
  X2 = runif(200),
  X3 = 1:20,
  x5 = 50:1
)

names(L1)

lapply(L1, mean)
```

## Importing some data

In the data directory there are some dummy text files: f1.csv, f2.txt and f3.txt
Import these sets in the ‘correct’ way. The resulting datasets in R should all have 3 columns.

```{r}
f1 <- read_csv("C:/Users/llam/OneDrive - RTL Nederland B.V/Personal/inleidingR/data/f1.csv",  skip = 3)

f2 <- read_delim("C:/Users/llam/OneDrive - RTL Nederland B.V/Personal/inleidingR/data/f2.txt",   " ")

f3 <- read.csv("C:/Users/llam/OneDrive - RTL Nederland B.V/Personal/inleidingR/data/f3.txt")
```

<br>

# Session 02. Data preparation

---

## String bewerkingen
Copy the following vector to R.
X = c("abd 06-456", "blab la (06)-123.45678", "hoi 06-12123478", "mijn nr 0689452312")

A.	For this exercise a valid mobile phone number starts with 06 followed by hyphen or space or not and then 8 consecutive digits. What are the valid numbers in x, use a regular expression.

B.	Extract the phone numbers from x into a new vector y

```{r}
library(stringr)
x = c("abd 06-456", "blab la (06)-123.45678", "hoi 06-12123478", "mijn nr 0689452312")

mob_pattern = "[0][6][[:punct:]]*[\\d]{8}"
str_detect(x,mob_pattern)

str_extract(x, mob_pattern)
```

## Dates and times.
A.	Create a date object from the characters "31-jan-05"   "22-Sep-07".

B.	Create a date object from the characters "31-jan-05 3:03:20"  "22-Sep-07 22:29:56".

C.	Add 4 months and 2 days to the dates from exercise 

D.	Create a date sequence vector x with the dates: 31-1-2016, 29-2-2016, 31-03-2016,……,31-12-2016 

```{r}
library(anytime)
x = anydate(c("31-jan-2005", "22-Sep-2007"))
y = anytime(c("31-jan-2005 3:03:20",  "22-Sep-2007 22:29:56"))

library(lubridate)

month(x) = month(x) +4
day(x) = day(x) +2
x

x = anydate( rep("2016-01-01", 12))
month(x) = 1:12
month(x) = month(x) + 1
x = x -1
```

For the following two exercises we are going to use restaurant data in the file Restaurants.csv in the data folder. Use the read_csv function to import this data set.

## Basic data manipulation 1
A.	Create a subset of the restaurants whose name is “kota radja”, how many are there?

B.	Give a top ten restaurants in Amsterdam based on price.

C.	Create an extra column that has the values zero or one.   1 if Kitchen type is Chinees and 0 otherwise.

D.	Create a sub set of the data, filter only the restaurants that have price smaller than 25.

E.	Give the number per kitchen types of that sub set.

```{r}
library(readr)
library(stringr)
library(dplyr)
restaurants = read_csv("data/Restaurants.csv")

restaurants %>% filter(
  str_detect(restNamen, "Kota")
  )

restaurants %>% filter( plaats == "Amsterdam") %>% arrange(desc(prijs)) %>% slice(1:10)

restaurants = restaurants %>% 
  mutate(
    Chinese = ifelse(typerest == "Chinees",1,0)
  )

elcheapo = restaurants %>% filter(prijs < 25)

elcheapo %>% group_by(keuken) %>% summarise(n=n())

```

## Basic data manipulation 2
A.	In the data folder there is a Postcode_NL.RDs file, import that set. 

B.	Use it to enrich the restaurant data set whith an extra column provincie.

C.	Count the number of restaurants per province of the whole set.

D.	Give the max and min prices per province.

E.	In the column “keuken” there are some mistakes. A price where there should have been a kitchen type. Remove all kitchen types that have a digit in.

```{r}
postcode = readRDS("data/postocdes_NL.RDs")
restaurants = left_join(restaurants, postcode, by=c("PCs"="Postcode_spatie"))

restaurants %>% 
  group_by(province) %>%
  summarise(n=n())

## pas op NA
restaurants %>% 
  group_by(province) %>%
  summarise(
    maxPrijs = max(prijs),
    minPrijs = min(prijs)
    )

restaurants %>% 
  group_by(province) %>%
  summarise(
    maxPrijs = max(prijs, na.rm = TRUE),
    minPrijs = min(prijs, na.rm = TRUE)
    )

restaurants = restaurants %>% 
  filter(
    !str_detect(keuken,"[:digit:]")
  )
```

## Auto data van gaspedaal
In the data folder there is a file R data set AllCarsGasPedaal.Rds  import this data set using `readRDS`.

A.	List the names of the columns of the data set.

B.	In this data set there is a column bjkm which is character, containing the kilometerstand, extract it and convert it to numeric so that we can calculate with it.

C.	Select only the brands Audi, Renault and BMW put those records in a new data frame.

```{r}
gaspedaal = readRDS("data/AllCarsGasPedaal.Rds")
names(gaspedaal)

### in stapjes
# HAAL EERST ALLES WAT VOOR KM STAAT
gaspedaal$tmp = str_extract(gaspedaal$bjkm, "[:digit:]*[\\.]*[:digit:]*\\skm")

#HAAL KM WEG EN HAAL PUNT WEG EN TRANSFORMEER NAAR NUMERIC
gaspedaal$KM = gaspedaal$tmp %>% 
  str_replace("km","") %>% 
  str_replace_all("\\.","") %>%
  as.numeric()

TEST = gaspedaal %>% filter(Merk %in% c("Audi","Renault","BMW"))

```

## Split data and stack data
Consider the following simple data set with 3 rows:

Df = data.frame(
  V1 = c(“a,b,r”, “p,q”, “p,q,w,z”)
)

I need the following set:
Id Item
1 a
1 b
1 r
2 p
2 q
3 p
3 q
3 w
3 z
How to this? use tidyr functions separate and gather.

```{r}
library(tidyverse)
Df = data.frame(
  V1 = c("a,b,r", "p,q", "p,q,w,z")
)

tmp = tidyr::separate(Df, V1, into = paste0("V",1:10)) %>%
  mutate(tmp = 1:length(V1)) %>%
  gather(id, item, -tmp) %>%
  filter(!is.na(item))
```

## sampling van data 
Kijk naar sample_n of sample_frac uit dplyr.

A. Create a df1 data frame with 100 columns and 1000 rows. Filled with random numbers between 0 and 1.

B. Create df2 which samples 500 rows from df1

C. Create df3 which samples 50% of the rows from df1

```{r}
# maak een 0/1 vector stop in een matrix en dan in een dataframe 
sample(c(0,1), replace = T, size = 100*1000) %>% matrix(ncol=100) %>% as.data.frame() -> df1

df2 = sample_n(df1,500)

df3 = sample_frac(df1, size=0.5)
```



<br>

# Sessie 03. Data visualisation

---

Use the set AllCarsGasPedaal.Rds  from the previous exercise.
```{r}
library(ggplot2)
library(dplyr)

AllCars = readRDS("data/AllCarsGasPedaal.Rds")
```

## plotting using ggplot2

A.	Create a scatterplot of the VraagPrijs and KMStand.<br>
B.	Remove outlying points (say KM > 500000 and Prijs > 100000) and create plot again using alpha = 0.05 <br>

D.	Filter the set so that we only have the top eight brands (based on # observations). <br>
E.	Create a new scatter plot, now the points should be colored  by brand, using the subset of the previous exercise. <br>

G.	Create facets by brand, each facet should have a histogram of the price, create also a grid of 2 rows and 4 columns. <br>
H.	Create facets by brand, each facet should contain a scatter plot between price and KM stand.


```{r}
AllCars = readRDS("data/AllCarsGasPedaal.Rds")

AllCars %>% 
  ggplot(aes(x=KMStand, y = VraagPrijs)) +
  geom_point()

AllCars %>% 
  dplyr::filter(VraagPrijs < 100000, KMStand < 500000) %>%
  ggplot(aes(x = KMStand, y = VraagPrijs)) +
  geom_point(alpha = 0.05)



top8 = AllCars %>%
  group_by(Merk) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n)) %>% 
  slice(1:8) %>% 
  select(Merk)

AllCars %>% 
  dplyr::filter(VraagPrijs < 100000, KMStand < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(x=KMStand, y = VraagPrijs)) +
  geom_point(aes(colour=Merk),alpha = 0.05) 



AllCars %>% 
  dplyr::filter(VraagPrijs < 100000, KMStand < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(VraagPrijs)) +
  geom_histogram() + facet_grid(~Merk)

AllCars %>% 
  dplyr::filter(VraagPrijs < 100000, KMStand < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(VraagPrijs)) +
  geom_histogram() + facet_wrap(~Merk, nrow=2)

AllCars %>% 
  dplyr::filter(VraagPrijs < 100000, KMStand < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(x= KMStand, y = VraagPrijs)) +
  geom_point() + facet_grid(~Merk)


```

## Smoothing plots

Let’s use the AllCarsGasPedaal again for continuous X and Y.

A.	Create a scatter plot of VraagPrijs and KMStand again but now add a smoothed line through the points to get an idea of the relation between KM and Price. <br>
B.	Also create a smoothing line per brand, only the top 8 brands. <br>

Let’s use the titanic_train data set in the titanic package for continuous X and binary Y.

C.	Create a scatter plot of Age and Survived. Not much interesting to see right? <br>
D.	Now add a smoothing line. Interesting relation. <br>
E.	Create facets for the variable Sex. <br>
F.	Create facets for the variables Sex and Pclass. <br>


```{r}
AllCars %>%
  ggplot(aes(x = KMStand, y = VraagPrijs)) +
  geom_point(alpha = 0.05) + 
  geom_smooth()

AllCars %>% 
  dplyr::filter(VraagPrijs < 100000, KMStand < 500000) %>%
  inner_join(top8) %>%
  ggplot(aes(x=KMStand, y = VraagPrijs)) +
  geom_point(aes(colour=Merk),alpha = 0.05) + 
  geom_smooth(aes(colour=Merk))


titanic::titanic_train %>%
  ggplot(aes(x=Age, y = Survived)) +
  geom_point()


titanic::titanic_train %>%
  ggplot(aes(x=Age, y = Survived)) +
  geom_point() + geom_smooth()


titanic::titanic_train %>%
  ggplot(aes(x=Age, y = Survived)) +
  geom_point() + geom_smooth() +
  facet_grid(~Sex)

titanic::titanic_train %>%
  ggplot(aes(x=Age, y = Survived)) +
  geom_point() + geom_smooth() +
  facet_grid(Sex~Pclass)
```


## Plotly

With plotly you can create interactive graphs. Lets use the AllCarsGasPedaal data again.

A.	Creating interactive scatter plots with 435K data points makes the plot very slow, first sample randomly 10% from the data and then plot a scatter with plotly KMStand against VraagPrijs. <br>
B.	Now create the same plot but when you hover over the points it should display name of the car, in the variable ‘auto’. <br>
C.	Create also a plot where the points are colored according to age. You need to calculate age first. <br>
D.	There are some outlying ages and prices, remove them so that the plot becomes better. <br>
E.	Create per Brand (Merk) a boxplot of the Price, because there are too many brands, focus on the top 25 brands (based on counts in the data set. Also remove outlying prices (price > 250K). <br>


```{r}
library(plotly)
library(dplyr)

AllCars %>%
  sample_frac(size=0.1) %>%
  plot_ly(x=~VraagPrijs, y = ~KMStand)

AllCars %>%
  sample_frac(size=0.1) %>%
  plot_ly(x=~VraagPrijs, y = ~KMStand, text = ~auto)


AllCars %>%
  mutate(leeftijd = 2017 - bouwjaar) %>%
  sample_frac(size=0.1) %>%
  plot_ly(x=~VraagPrijs, y = ~KMStand, text = ~auto, color = ~leeftijd)


AllCars %>%
  mutate(leeftijd = 2017 - bouwjaar) %>%
  filter(leeftijd < 15, VraagPrijs < 250000) %>%
  sample_frac(size=0.1) %>%
  plot_ly(x=~VraagPrijs, y = ~KMStand, text = ~auto, color = ~leeftijd)



top25 = AllCars %>%
  group_by(Merk) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n)) %>% 
  slice(1:25) %>% 
  select(Merk)

AllCars %>% 
  filter(VraagPrijs < 250000) %>%
  inner_join(top25) %>%
  plot_ly( y = ~VraagPrijs, color = ~Merk,type = "box")
```



## Leaflets

```{r}
library(leaflet)
library(dplyr)

Restaurants <- read_csv("data/Restaurants.csv")

Utrecht = Restaurants %>% filter(plaats == "Utrecht")

m2 = leaflet(data = Utrecht) %>%
  addTiles() %>%  
  addMarkers(lng = ~LONGs, lat = ~LATs)
m2  


m2 = Restaurants %>% 
  filter(
    plaats == "Utrecht",
    LATs > 50
  ) %>%
  leaflet() %>%
  addTiles() %>%  
  addMarkers(lng = ~LONGs, lat = ~LATs)
m2  



m2 = Restaurants %>% 
  filter(
    plaats == "Utrecht",
    LATs > 50
  ) %>%
  leaflet() %>%
  addTiles() %>%  
  addMarkers(lng = ~LONGs, lat = ~LATs, popup = ~restNamen)
m2  
```



## Leaflets with polygons

Instead of putting markers or circles on a leaflet, polygons can also be placed on an interactive map. We use the provinces of the Netherlands. Data is already present in the shape file TopGrenzen-prov-actueel.shp. <br>

A.	Import the TopGrenzen shape file as in the lecture. <br>
B.	Plot the polygons of the Dutch provinces on a leaflet with the function addPolygons. <br>

The polygons can have different colors, for each province we can calculate the average prijs for a menu in the restaurant, and color the polygon accordingly. First calculate the average prijs per province. <br>

C.	To do that we need to tell in which province a restaurant is, import the data set postcodes_NL.RDs in the data directory. <br>
D.	Join the restaurant data with this postcode data set on postocde so that we know for each restaurant in which province they are. <br>
E.	Now group by province and calculate per province the average prijs (be aware of NA prices!!). <br>
F.	Join the prov polygons data with the average prices. <br>
G.	Define a four color scale for the price with the function. <br> 
H.	Now create a leaflet and use addPolygons to draw per province the price. <br>
I.	In addPolygons you can specify highlight options, experiment with that. <br>


```{r}
library(maptools)
library(colorRamps)

ProvNL <- readShapeSpatial("data/TopGrenzen-prov-actueel.shp")
proj4string(ProvNL) <- CRS("+init=epsg:28992")
ProvNL = spTransform(ProvNL, CRS("+proj=longlat +datum=WGS84"))


leaflet(ProvNL) %>%
  addTiles() %>%
  addPolygons(
    stroke = TRUE, weight = 1, fillOpacity = 0.15, smoothFactor = 0.15
  )


## import postcode om te matchen met restaurants
postocdes_NL <- readRDS("data/postocdes_NL.RDs")
Restaurants = Restaurants %>% left_join(postocdes_NL, by = c("PCs"= "Postcode_spatie")) 

## nu kan je per provincie de gemiddelde prijs uitrekenen
Prijs = Restaurants %>% group_by(province) %>% summarise(prijs = mean(prijs, na.rm = TRUE))

## de ProvNL shape data, haal de data set er uit en match corresponerende gemiddelde prijs per provincie toe
TMP = ProvNL@data
TMP$province = as.character(TMP$Provincien)
TMP = TMP %>% left_join(Prijs)

## Zet de gemodificeerde data weer terug
ProvNL@data = TMP

### definieer een 4 niveau schema op basis van de prijs
colpal <- colorQuantile(
  palette = green2red(4), n=4,
  domain = ProvNL$prijs
)

### teken nu de leaflet weer
leaflet(ProvNL) %>%
  addTiles() %>%
  addPolygons(
    stroke = TRUE, weight = 1, 
    fillOpacity = 0.25, smoothFactor = 0.15,
    fillColor = colpal(ProvNL$prijs)
  )


### teken nu nog een leaflet met highlight opties
leaflet(ProvNL) %>%
  addTiles() %>%
  addPolygons(
    stroke = TRUE, weight = 1,
    fillOpacity = 0.25, smoothFactor = 0.15, 
    fillColor = colpal(ProvNL$prijs) ,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    )
  )


### teken nu nog een leaflet met popup texts

labels <- sprintf(
  "<strong>%s</strong><br/>  gemiddelde prijs %g ",
  ProvNL$province, ProvNL$prijs
) %>% lapply(htmltools::HTML)

leaflet(ProvNL) %>%
  addTiles() %>%
  addPolygons(
    stroke = TRUE, weight = 1,
    fillOpacity = 0.25, smoothFactor = 0.15, 
    fillColor = colpal(ProvNL$prijs) ,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")
  ) %>%
  addLegend(  
    pal = colpal, values = ProvNL$prijs,
    opacity = 0.7, title = NULL,
    position = "bottomright"
  )





```



# Session 4. Machine learning

## Predicting the price in the Cars data

```{r}
library(dplyr)
library(splines)
library(forcats)

AllCars = readRDS("data/AllCarsGasPedaal.Rds")

### simple linear regression
model1 = lm(VraagPrijs ~ KMStand, data = AllCars)
summary(model1)

### buckets in KMStand, lets create 15 buckets in in the KMStand variable
AllCars = AllCars %>% mutate(
  KMBuckets = cut(KMStand, 15)
)

model2 = lm(VraagPrijs ~ KMBuckets, data = AllCars)
summary(model2)

### spline model with Natural Splines
model3 = lm(VraagPrijs ~ ns(KMStand,5), data = AllCars)
summary(model3)

### More variables in the model
table(AllCars$Merk)

AllCars = AllCars %>% 
  mutate(
    Merk2 = fct_lump(Merk,10)
  )

model4 = lm(
  VraagPrijs ~
    ns(KMStand,5) + 
    Merk2 + 
    ns(OuderdomMaanden,5) + 
    Brandstof + 
    Transmissie ,
  data = AllCars
)

summary(model4)
```

## Using random forests to predit the price

```{r}
library(h2o)
library(dplyr)

h2o.init()

## POSIXct type not supported in h2o, remove column
AllCars$Bouwdatum = NULL
carh2o = as.h2o(AllCars)
h2o.ls()
m1 = h2o.randomForest(
  x = "KMStand",
  y = "VraagPrijs", 
  training_frame = carh2o,
  validation_frame = carh2o
)

m1

## predict on carh2o data set and retrieve to R
newcars = predict(m1, newdata = carh2o) %>% as.data.frame()
AllCars = cbind(AllCars, newcars)

## calculating R2 your self with a prediction and target
R2 = 1 - sum((AllCars$VraagPrijs  - AllCars$predict)^2) / sum( (AllCars$VraagPrijs - mean(AllCars$VraagPrijs))^2) 
```






